### 2017-11-29 submission/manuscript.Rmd
re write discussion and conclusion

The data presented herein supports the driver-passenger model of the microbial role in CRC, as summarized by Flynn [@flynn_metabolic_2016], when applied to carcinoma but not necessarily adenoma. Both the drastically increased RR of CRC-associated genera versus diversity for carcinoma and increasing RR with more CRC-associated generea positivity are highly supportive of this model. It is also possible that in a driver-passenger scenario, simply having the driver present or only identifying the passenger is a good enough proxy that the event is occuring. This would account for the observation that there is no constant additive effect on RR for increasing positivity. Additionally, the initial establishment of the driver within the system appears to be dependent on the current community. This is supported by our finding that when adding the community context to our models in addition to the CRC-associated genera, the model AUC increases.

With regards to carcinoma, our results support the driver-passenger model within the framework of the transition from adenoma to carcinoma. Conversely, using the present data, we can only theorize observations related to adenoma development fit this model. The changes that occur at this timepoint are small and possibly focal to the adenoma itself. The stepwise decrease in diversity suggests that the adenoma community is not normal but has changed subtly. Although there appears to be localized changes that do depend on the driver-passenger model, as supported by an increased RR for one, two, and four positive CRC-associated genera in tissue [Figure 3C], there may be other processes at play that ultimately exacerbate the condition from a subtle localized change to a change in the global community. The poor performance of the Random Forest models for classifying adenoma based only on the microbiota would suggest that this is the case. It is possible to hypothesize that at early stages of the diease, the host interacts with these subtle changes and catalyzes what ultimately leads to a thoroughly dysfunctional community.

Although there are still questions that need to be answered regarding the microbiota and carcinoma, a clearer framework for their relationship is beginning to develop as to how this occurs. From our observations many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control and most studies analyzed have sufficient power to detect these types of changes [Figure 6]. Conversely, the role of the microbiota in adenoma is less clear and part of the reason this may be is because many studies are not powered effectively to observe the small changes reported here. None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls. This small effect size range may well be the scope in which differences consistently occur in adenoma due to the subtle changes in community that occur between control and adenoma. Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out how the microbiota contributes to adenoma formation.

@@

By aggregating together a large collection of studies from both feces and tissue, we are able to provide evidence in support of the importance of the bacterial community in both adenoma and carcinoma. We are also able to provide support for the driver-passenger model in the context of carcinoma. However, within the context of adenoma, it is less clear that this relationship exists. These observations highlight the importance of power and sample number considerations when undertaking investigations into the microbiota and adenoma due to the subtle changes in the community. Although there are power limitations associated with adenoma, this report highlights the strong influence the microbiota has on CRC development.

>>>>

The data presented herein supports the driver-passenger model of the microbial role in CRC, as summarized by Flynn [@flynn_metabolic_2016], when applied to carcinoma but not necessarily adenoma. The central idea of the model is that a single bacterium initiates an environment in which other non-resident microbes may then be able to colonize, creating a vicious cycle that is conducive for CRC. Both the drastically increased RR of CRC-associated genera versus $\alpha$-diversity metrics for carcinoma and increasing RR with more CRC-associated generea positivity are highly supportive of this model. However, the initial establishment of the driver within the system appears to be dependent on the current community. This is supported by our finding that when adding the community context to our models in addition to the CRC-associated genera, the model AUC increases [Figure 4 & 5]. Conversely, using the present data, it is less likely that adenoma development fits this model. The changes that occur at this timepoint are small and possibly focal to the adenoma itself. The stepwise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1]. Although there appears to be localized changes that do depend on the driver-passenger model, as supported by an increased RR for one, two, and four positive CRC-associated genera in tissue [Figure 3C], there may be other processes at play that ultimately exacerbate the condition from a subtle localized change, to a change in the global community. The poor performance of the Random Forest models for classifying adenoma based only on the microbiota would suggest that this is the case. One potential hypothesis from these observations is that at early stages of the diease, how the host interacts with these subtle changes is what ultimately leads to a thoroughly dysfunctional community that is supportive of CRC genesis.

Within stool, common resident microbes were most consistently present in the top 10 genera or OTUs across study [Figure 6]. Changes in *Bacteroides*, *Ruminococcus*, and *Roseburia* were consistently found to be disciminative across the different studies for both adenoma and carcinoma models [Figure 6]. This data would suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* is not as important as how these bacteria interact with the changing resident community. These observations would also suggest that initial changes in the resident community, specifically to *Ruminococcaceae*, *Ruminococcus*, and *Roseburia*, carry on from adenoma to carcinoma. Based on these observations, the early changes in the community may give rise to initial polyp formation via interactions with the host. These changes then create new niches in which any one of the CRC-associated genera could gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma via a driver-passenger type mechanism.

The tissue studies did not provide a clearer understanding of how the microbiota may drive the progression of disease severity. For the OTU models both the unmatched and matched [Figure S7E & F] tissue samples had some concordance with the stool data, with resident bacteria being the most prevalent in the top 10 across studies. Unlike in stool, *Fusobacterium* was the only CRC-associated bacteria consistently present in the top 10 of the CRC models [Figure S7B-C & E-F]. The majority of the results seem to be study specific with many top 10 taxa being present only in a single study. One other potentially worrying sign is the presence of *Propionibacterium* within the top 10 for the genera and OTU models and could be a marker of contamination. The low biomass of these samples coupled with potential contamination might be a possible reason why the tissue results seem to be more sporadic then the stool results.

Although there are still questions that need to be answered regarding the microbiota and carcinoma, a clearer framework for their relationship is beginning to develop as to how this occurs. From our observations many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control and most studies analyzed have sufficient power to detect these types of changes [Figure 7]. Conversely, the role of the microbiota in adenoma is less clear and part of the reason this may be is because many studies are not powered effectively to observe the small changes reported here. None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls. This small effect size range may well be the scope in which differences consistently occur in adenoma due to the subtle changes in community that occur between control and adenoma. Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out how the microbiota contributes to adenoma formation.

@@

By aggregating together a large collection of studies from both feces and tissue, we are able to provide evidence in support of the importance of the bacterial community in both adenoma and carcinoma. Overall, our results support a framework by which early localized community changes give rise to polyp formation. With the host as a potential catalyst, new niches arise by which non-resident CRC-associated microbes can then gain a foothold and create an environment that allows more of these microbes to colonize. This exacerbates the existing community changes and creates a vicious cycle conducive of carcinoma formation. Our observations also highlight the importance of power and sample number considerations when undertaking investigations into the microbiota and adenoma due to the subtle changes in the community. Although there are power limitations associated with adenoma, this report highlights the strong influence the microbiota has on CRC development.

### 2017-11-29 submission/manuscript.tex
re write discussion and conclusion

The data presented herein supports the driver-passenger model of the
microbial role in CRC, as summarized by Flynn {[}2{]}, when applied to
carcinoma but not necessarily adenoma. Both the drastically increased RR
of CRC-associated genera versus diversity for carcinoma and increasing
RR with more CRC-associated generea positivity are highly supportive of
this model. It is also possible that in a driver-passenger scenario,
simply having the driver present or only identifying the passenger is a
good enough proxy that the event is occuring. This would account for the
observation that there is no constant additive effect on RR for
increasing positivity. Additionally, the initial establishment of the
driver within the system appears to be dependent on the current
community. This is supported by our finding that when adding the
community context to our models in addition to the CRC-associated
genera, the model AUC increases.

With regards to carcinoma, our results support the driver-passenger
model within the framework of the transition from adenoma to carcinoma.
Conversely, using the present data, we can only theorize observations
related to adenoma development fit this model. The changes that occur at
this timepoint are small and possibly focal to the adenoma itself. The
stepwise decrease in diversity suggests that the adenoma community is
not normal but has changed subtly. Although there appears to be
localized changes that do depend on the driver-passenger model, as
supported by an increased RR for one, two, and four positive
CRC-associated genera in tissue {[}Figure 3C{]}, there may be other
processes at play that ultimately exacerbate the condition from a subtle
localized change to a change in the global community. The poor
performance of the Random Forest models for classifying adenoma based
only on the microbiota would suggest that this is the case. It is
possible to hypothesize that at early stages of the diease, the host
interacts with these subtle changes and catalyzes what ultimately leads
to a thoroughly dysfunctional community.

Although there are still questions that need to be answered regarding
the microbiota and carcinoma, a clearer framework for their relationship
is beginning to develop as to how this occurs. From our observations
many changes in carcinoma could easily result in effect sizes that are
30\% or more between the case and control and most studies analyzed have
sufficient power to detect these types of changes {[}Figure 6{]}.
Conversely, the role of the microbiota in adenoma is less clear and part
of the reason this may be is because many studies are not powered
effectively to observe the small changes reported here. None of the
studies analyzed were properly powered to detect a 10\% or lower change
between case and controls. This small effect size range may well be the
scope in which differences consistently occur in adenoma due to the
subtle changes in community that occur between control and adenoma.
Future studies investigating adenoma and the microbiota need to take
these factors into consideration if we are to work out how the
microbiota contributes to adenoma formation.

>>>>

The data presented herein supports the driver-passenger model of the
microbial role in CRC, as summarized by Flynn {[}2{]}, when applied to
carcinoma but not necessarily adenoma. The central idea of the model is
that a single bacterium initiates an environment in which other
non-resident microbes may then be able to colonize, creating a vicious
cycle that is conducive for CRC. Both the drastically increased RR of
CRC-associated genera versus \(\alpha\)-diversity metrics for carcinoma
and increasing RR with more CRC-associated generea positivity are highly
supportive of this model. However, the initial establishment of the
driver within the system appears to be dependent on the current
community. This is supported by our finding that when adding the
community context to our models in addition to the CRC-associated
genera, the model AUC increases {[}Figure 4 \& 5{]}. Conversely, using
the present data, it is less likely that adenoma development fits this
model. The changes that occur at this timepoint are small and possibly
focal to the adenoma itself. The stepwise decrease in diversity suggests
that the adenoma community is not normal but has changed subtly
{[}Figure 1{]}. Although there appears to be localized changes that do
depend on the driver-passenger model, as supported by an increased RR
for one, two, and four positive CRC-associated genera in tissue
{[}Figure 3C{]}, there may be other processes at play that ultimately
exacerbate the condition from a subtle localized change, to a change in
the global community. The poor performance of the Random Forest models
for classifying adenoma based only on the microbiota would suggest that
this is the case. One potential hypothesis from these observations is
that at early stages of the diease, how the host interacts with these
subtle changes is what ultimately leads to a thoroughly dysfunctional
community that is supportive of CRC genesis.

Within stool, common resident microbes were most consistently present in
the top 10 genera or OTUs across study {[}Figure 6{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcus}, and \emph{Roseburia} were
consistently found to be disciminative across the different studies for
both adenoma and carcinoma models {[}Figure 6{]}. This data would
suggest that whether the non-resident bacterium is \emph{Fusobacteria}
or \emph{Peptostreptococcus} is not as important as how these bacteria
interact with the changing resident community. These observations would
also suggest that initial changes in the resident community,
specifically to \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
\emph{Roseburia}, carry on from adenoma to carcinoma. Based on these
observations, the early changes in the community may give rise to
initial polyp formation via interactions with the host. These changes
then create new niches in which any one of the CRC-associated genera
could gain a foothold, exacerbating the initial changes in community and
facilitating the transition from adenoma to carcinoma via a
driver-passenger type mechanism.

The tissue studies did not provide a clearer understanding of how the
microbiota may drive the progression of disease severity. For the OTU
models both the unmatched and matched {[}Figure S7E \& F{]} tissue
samples had some concordance with the stool data, with resident bacteria
being the most prevalent in the top 10 across studies. Unlike in stool,
\emph{Fusobacterium} was the only CRC-associated bacteria consistently
present in the top 10 of the CRC models {[}Figure S7B-C \& E-F{]}. The
majority of the results seem to be study specific with many top 10 taxa
being present only in a single study. One other potentially worrying
sign is the presence of \emph{Propionibacterium} within the top 10 for
the genera and OTU models and could be a marker of contamination. The
low biomass of these samples coupled with potential contamination might
be a possible reason why the tissue results seem to be more sporadic
then the stool results.

Although there are still questions that need to be answered regarding
the microbiota and carcinoma, a clearer framework for their relationship
is beginning to develop as to how this occurs. From our observations
many changes in carcinoma could easily result in effect sizes that are
30\% or more between the case and control and most studies analyzed have
sufficient power to detect these types of changes {[}Figure 7{]}.
Conversely, the role of the microbiota in adenoma is less clear and part
of the reason this may be is because many studies are not powered
effectively to observe the small changes reported here. None of the
studies analyzed were properly powered to detect a 10\% or lower change
between case and controls. This small effect size range may well be the
scope in which differences consistently occur in adenoma due to the
subtle changes in community that occur between control and adenoma.
Future studies investigating adenoma and the microbiota need to take
these factors into consideration if we are to work out how the
microbiota contributes to adenoma formation.
